---
title: "Module 03: Git, SQL, and the OMOP CDM"
subtitle: "Hands-on Practice with Version Control and Healthcare Data"
---

```{r}
#| include: false
# We're not using scaffold() here since this is a teaching module
# about the fundamentals that scaffold() abstracts away.
```

## Agenda

1. **Git Practice** (Hands-on)
2. **OMOP CDM Schema Tour** (Exploration)
3. **Eunomia Setup & SQL Basics** (Hands-on)
4. **SqlRender Across Databases** (Demo)
5. **Patient Odyssey** (Guided Exploration)
6. **ATLAS Cohort Definitions** (Exploration)

---

## 1. Git Practice

We'll create a Git repository using RStudio, practice the workflow with both the terminal and RStudio's built-in tools.

### Create a new RStudio Project

1. In RStudio: **File -> New Project...**
2. Select **New Directory**
3. Select **New Project**
4. Name it `git-practice`
5. Choose a location (e.g., your Desktop)
6. Click **Create Project**

RStudio opens your new project. Notice there's no Git pane yet.

*(Note: RStudio has a "Create a git repository" checkbox you can use to skip the next step, but we'll do it manually to learn the command.)*

### Initialize Git via terminal

Open the Terminal in RStudio (Tools -> Terminal -> New Terminal).

```bash
git init
```

You should see: `Initialized empty Git repository in .../git-practice/.git/`

Now look at RStudio. The **Git pane** should appear (usually top-right). If not, restart RStudio or go to Tools -> Project Options -> Git/SVN and confirm Git is detected.

### Copy this module file

Copy this module file (`ohdsi-basics.qmd`) into your new `git-practice` folder. You can:

- Use your file browser to copy/paste, or
- In the RStudio Files pane, click **More -> Set As Working Directory**, then copy the file

### Explore RStudio's Git pane

Look at the **Git pane** (usually top-right, next to Environment/History).

You should see `ohdsi-basics.qmd` listed with a yellow **?** icon. This means it's untracked.

### Stage using RStudio

In the Git pane:

1. Check the box next to `ohdsi-basics.qmd`
2. The icon changes to a green **A** (Added/staged)

That's it. You just staged a file without touching the terminal.

### Commit using RStudio

1. Click **Commit** in the Git pane
2. A new window opens showing your staged changes
3. Write a commit message: "Add module file"
4. Click **Commit**

You've made your first commit.

### Now let's use the Terminal

RStudio's GUI is convenient, but knowing the terminal commands matters. Open the Terminal (Tools -> Terminal -> New Terminal).

```bash
# Check status
git status
```

Git shows a clean working directory. Your commit worked.

```bash
# View history
git log --oneline
```

You'll see your commit with its message.

### Make changes and commit via terminal

Edit the module file: add a comment at the top, or change something small. Save it.

```bash
# See what changed
git status
git diff
```

`git diff` shows the exact lines that changed.

```bash
# Stage the file
git add ohdsi-basics.qmd

# Commit with a message
git commit -m "Make edits to module"
```

### Compare: RStudio vs Terminal

- **See changes:** Git pane shows icons, or `git status`
- **Stage:** Check the box, or `git add <file>`
- **Commit:** Click Commit button, or `git commit -m "msg"`
- **View diff:** Click file in Commit window, or `git diff`
- **History:** Click History, or `git log`

Both work. Use whichever feels natural.

### Key terminal commands

- `git status`: See current state
- `git add <file>`: Stage changes
- `git commit -m "msg"`: Save snapshot
- `git diff`: See unstaged changes
- `git log`: View history

### Connecting to GitHub

When you're ready to backup to GitHub:

1. Go to GitHub and create a new repository (don't initialize with README)
2. Copy the repository URL

In the terminal:

```bash
# Connect your local repo to GitHub
git remote add origin https://github.com/YOUR-USERNAME/git-practice.git

# Push your commits
git push -u origin main
```

Or in RStudio: **Tools -> Project Options -> Git/SVN** and add the remote URL.

After the first push, you can use:

- Terminal: `git push` and `git pull`
- RStudio: Push/Pull buttons in the Git pane

### .gitignore

Not everything belongs in Git. A `.gitignore` file tells Git which files to skip.

Common things to ignore:

- Large data files (`*.csv`, `*.rds`, `data/`)
- Credentials and secrets (`.env`, `.Renviron`, `*.pem`)
- Generated outputs (`*_files/`, `*.html`, `_site/`)
- System junk (`.DS_Store`, `Thumbs.db`)

Create a `.gitignore` file in your project root:

```
# Data files
data/
*.csv
*.rds

# Secrets
.env
.Renviron

# Generated output
*_files/
*.html

# System
.DS_Store
```

RStudio projects create a default `.gitignore` for you. Open it and review what's already there.

**Why this matters:** Without `.gitignore`, you might accidentally commit a 500MB data file or push database credentials to public GitHub.

**Critical for biomedical work:** Patient data and PHI should never be committed to a Git repository. Once something is committed, it lives in the history forever, even if you delete the file later. Even "private" repositories can be misconfigured or leaked. Always ignore your `data/` folder and query data from databases rather than storing local copies.

### Fetch vs Pull

```bash
git fetch   # Download changes from GitHub (don't merge yet)
git pull    # Download AND merge changes
```

`fetch` lets you see what's new before merging. `pull` does both at once.

### Clean up (optional)

When you're done practicing, you can delete the `git-practice` folder from your Desktop.

---

## 2. OMOP CDM Schema Tour

The OMOP Common Data Model defines the structure for healthcare data. Let's explore the official specification.

### Browse the CDM Repository

Open in your browser: <https://github.com/OHDSI/CommonDataModel>

**Exploration tasks:**

1. Navigate to `inst/ddl/5.4/` - these are the table definitions
2. Open `postgresql/` and look at `OMOPCDM_postgresql_5.4_ddl.sql`
3. Find the `CREATE TABLE person` statement
4. What columns does the `person` table have?

### CDM Table Relationships

The CDM tables connect through `person_id` and `concept_id`.

**Core clinical tables** (all link to `person` via `person_id`):

- `person`: Demographics (birth year, gender, race)
- `observation_period`: When we have data on each person
- `visit_occurrence`: Healthcare encounters
- `condition_occurrence`: Diagnoses
- `drug_exposure`: Medications
- `procedure_occurrence`: Procedures performed
- `measurement`: Labs and vitals

**The vocabulary table:**

- `concept`: Standardized meanings for all coded values

Clinical events link to visits via `visit_occurrence_id`, and everything uses `concept_id` to reference standardized terminology.

### Key insight

Everything connects through:

- `person_id` -- links clinical events to a patient
- `concept_id` -- standardizes meaning across vocabularies

---

## 3. Eunomia Setup & SQL Basics

Eunomia is a synthetic OMOP CDM database we can query locally.

### Install packages (if needed)

```{r}
#| eval: false
install.packages("DatabaseConnector")
install.packages("Eunomia")
install.packages("SqlRender")
```

### Connect to Eunomia

```{r}
#| eval: false
library(DatabaseConnector)
library(Eunomia)

# Get connection details (downloads data on first run)
connectionDetails <- getEunomiaConnectionDetails()

# Connect
connection <- connect(connectionDetails)
```

### See what's in the database

```{r}
#| eval: false
# List all tables
getTableNames(connection, databaseSchema = "main")
```

### Query the person table

```{r}
#| eval: false
querySql(connection, "
  SELECT *
  FROM main.person
  LIMIT 10
")
```

Notice the `concept_id` columns - these are codes, not human-readable names.

### Join with concept to decode

```{r}
#| eval: false
querySql(connection, "
  SELECT
    p.person_id,
    p.year_of_birth,
    g.concept_name AS gender,
    r.concept_name AS race
  FROM main.person p
  LEFT JOIN main.concept g ON p.gender_concept_id = g.concept_id
  LEFT JOIN main.concept r ON p.race_concept_id = r.concept_id
  LIMIT 20
")
```

### Count patients by gender

```{r}
#| eval: false
querySql(connection, "
  SELECT
    g.concept_name AS gender,
    COUNT(*) AS n_patients
  FROM main.person p
  LEFT JOIN main.concept g ON p.gender_concept_id = g.concept_id
  GROUP BY g.concept_name
")
```

### Find the most common conditions

```{r}
#| eval: false
querySql(connection, "
  SELECT
    c.concept_name AS condition,
    COUNT(*) AS n_occurrences,
    COUNT(DISTINCT co.person_id) AS n_patients
  FROM main.condition_occurrence co
  JOIN main.concept c ON co.condition_concept_id = c.concept_id
  GROUP BY c.concept_name
  ORDER BY n_patients DESC
  LIMIT 15
")
```

### Find the most common drugs

```{r}
#| eval: false
querySql(connection, "
  SELECT
    c.concept_name AS drug,
    COUNT(*) AS n_exposures,
    COUNT(DISTINCT de.person_id) AS n_patients
  FROM main.drug_exposure de
  JOIN main.concept c ON de.drug_concept_id = c.concept_id
  GROUP BY c.concept_name
  ORDER BY n_patients DESC
  LIMIT 15
")
```

---

## 4. SqlRender Across Databases

OHDSI needs SQL that works on PostgreSQL, SQL Server, Oracle, and more. SqlRender translates.

### Load SqlRender

```{r}
#| eval: false
library(SqlRender)
```

### The problem: SQL dialects differ

```{r}
#| eval: false
# SQL Server syntax for "top N rows"
sql_server_style <- "SELECT TOP 10 * FROM person;"

# PostgreSQL uses LIMIT
postgresql_style <- "SELECT * FROM person LIMIT 10;"

# Oracle uses ROWNUM
oracle_style <- "SELECT * FROM person WHERE ROWNUM <= 10;"
```

Three ways to say the same thing!

### SqlRender translates for you

```{r}
#| eval: false
# Write in OHDSI SQL (SQL Server-ish syntax)
sql <- "SELECT TOP 10 * FROM person;"

# Translate to PostgreSQL
translate(sql, targetDialect = "postgresql")

# Translate to Oracle
translate(sql, targetDialect = "oracle")

# Translate to SQLite (what Eunomia uses)
translate(sql, targetDialect = "sqlite")
```

### Parameterization with @variables

```{r}
#| eval: false
# Template SQL with parameters
sql <- "
SELECT *
FROM @cdm_schema.person
WHERE year_of_birth >= @min_year
  AND year_of_birth <= @max_year
"

# Render with specific values
render(sql,
  cdm_schema = "main",
  min_year = 1950,
  max_year = 1970
)
```

### Combine render + translate

```{r}
#| eval: false
sql <- "
SELECT TOP 100 *
FROM @cdm_schema.condition_occurrence
WHERE condition_concept_id = @concept_id
"

# Render parameters
rendered <- render(sql,
  cdm_schema = "main",
  concept_id = 201826  # Type 2 diabetes
)

# Translate to target dialect
translated <- translate(rendered, targetDialect = "sqlite")

# Execute
querySql(connection, translated)
```

### Why this matters

You write SQL once. SqlRender handles the differences between:

- Your local SQLite (Eunomia)
- Your institution's SQL Server
- A collaborator's PostgreSQL
- A cloud Redshift cluster

This is how OHDSI runs the same study at 100+ sites.

---

## 5. Patient Odyssey

Let's follow a single patient through the CDM and understand their healthcare journey.

### Find a patient with interesting data

```{r}
#| eval: false
# Find patients with multiple conditions and drugs
interesting_patients <- querySql(connection, "
  SELECT
    p.person_id,
    COUNT(DISTINCT co.condition_concept_id) AS n_conditions,
    COUNT(DISTINCT de.drug_concept_id) AS n_drugs,
    COUNT(DISTINCT vo.visit_occurrence_id) AS n_visits
  FROM main.person p
  LEFT JOIN main.condition_occurrence co ON p.person_id = co.person_id
  LEFT JOIN main.drug_exposure de ON p.person_id = de.person_id
  LEFT JOIN main.visit_occurrence vo ON p.person_id = vo.person_id
  GROUP BY p.person_id
  HAVING n_conditions > 5 AND n_drugs > 5
  ORDER BY n_visits DESC
  LIMIT 10
")

interesting_patients
```

Pick a `person_id` from this list for our odyssey.

### Get patient demographics

```{r}
#| eval: false
# Replace with your chosen person_id
patient_id <- 1  # Use a real ID from above

querySql(connection, sprintf("
  SELECT
    p.person_id,
    p.year_of_birth,
    g.concept_name AS gender,
    r.concept_name AS race
  FROM main.person p
  LEFT JOIN main.concept g ON p.gender_concept_id = g.concept_id
  LEFT JOIN main.concept r ON p.race_concept_id = r.concept_id
  WHERE p.person_id = %d
", patient_id))
```

### Get observation period

When do we have data on this patient?

```{r}
#| eval: false
querySql(connection, sprintf("
  SELECT
    observation_period_start_date,
    observation_period_end_date
  FROM main.observation_period
  WHERE person_id = %d
", patient_id))
```

### Get their conditions

```{r}
#| eval: false
patient_conditions <- querySql(connection, sprintf("
  SELECT
    co.condition_start_date,
    c.concept_name AS condition,
    c.concept_id,
    v.concept_name AS visit_type
  FROM main.condition_occurrence co
  JOIN main.concept c ON co.condition_concept_id = c.concept_id
  LEFT JOIN main.visit_occurrence vo ON co.visit_occurrence_id = vo.visit_occurrence_id
  LEFT JOIN main.concept v ON vo.visit_concept_id = v.concept_id
  WHERE co.person_id = %d
  ORDER BY co.condition_start_date
", patient_id))

patient_conditions
```
### Get their drug exposures

```{r}
#| eval: false
patient_drugs <- querySql(connection, sprintf("
  SELECT
    de.drug_exposure_start_date,
    c.concept_name AS drug,
    c.concept_id,
    de.quantity,
    de.days_supply
  FROM main.drug_exposure de
  JOIN main.concept c ON de.drug_concept_id = c.concept_id
  WHERE de.person_id = %d
  ORDER BY de.drug_exposure_start_date
", patient_id))

patient_drugs
```

### Visualize the patient timeline

```{r}
#| eval: false
library(dplyr)
library(ggplot2)

# Combine conditions and drugs into a timeline
conditions_timeline <- patient_conditions |>
  select(date = CONDITION_START_DATE, event = CONDITION, type = "Condition")

drugs_timeline <- patient_drugs |>
  select(date = DRUG_EXPOSURE_START_DATE, event = DRUG, type = "Drug")

timeline <- bind_rows(conditions_timeline, drugs_timeline) |>
  mutate(date = as.Date(date)) |>
  arrange(date)

# Plot
ggplot(timeline, aes(x = date, y = event, color = type)) +
  geom_point(size = 3) +
  labs(
    title = sprintf("Patient %d Healthcare Timeline", patient_id),
    x = "Date",
    y = "",
    color = "Event Type"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))
```

### How the data connects

For any patient, the data flows like this:

1. Start with `person` (demographics)
2. Check `observation_period` to see when we have data
3. Look at `visit_occurrence` to find their encounters
4. From visits, find clinical events: `condition_occurrence`, `drug_exposure`, `procedure_occurrence`, `measurement`
5. Join any `*_concept_id` column to `concept` to get human-readable names

### Questions to explore

1. Does this patient have conditions that match their drug exposures? (e.g., diabetes + metformin)
2. What's the time gap between diagnosis and first treatment?
3. Are there any emergency visits? What happened?

---

## 6. ATLAS Cohort Definitions

ATLAS is OHDSI's web-based tool for defining cohorts and analyses. Let's see how cohorts translate to SQL.

### Access the ATLAS demo

Open in your browser: <https://atlas-demo.ohdsi.org>

*(Note: This is a public demo instance - no real patient data)*

### Explore existing cohort definitions

1. Click **Cohort Definitions** in the left menu
2. Browse the existing cohorts
3. Click on one to see its definition

### Anatomy of a cohort definition

A cohort has:

- **Entry criteria**: How does someone enter the cohort?
- **Inclusion criteria**: Additional requirements to stay in
- **Exit criteria**: When does someone leave the cohort?

### View the generated SQL

In any cohort definition:

1. Click the **Export** tab
2. Select **SQL** format
3. See the generated SQL code

This SQL can run on any OMOP CDM database!

### Example: Type 2 Diabetes cohort

A simple T2DM cohort might have:

**Entry criteria:**
- First diagnosis of Type 2 diabetes (concept 201826)

**Inclusion criteria:**
- Age ≥ 18 at diagnosis
- At least 365 days of prior observation

**Exit criteria:**
- End of observation period

### The generated SQL

ATLAS generates complex SQL that:

1. Finds all qualifying entry events
2. Applies inclusion criteria
3. Handles overlapping periods
4. Outputs the standard cohort format

```sql
-- Simplified example of what ATLAS generates
INSERT INTO cohort (
  cohort_definition_id,
  subject_id,
  cohort_start_date,
  cohort_end_date
)
SELECT
  1 AS cohort_definition_id,
  person_id AS subject_id,
  condition_start_date AS cohort_start_date,
  observation_period_end_date AS cohort_end_date
FROM condition_occurrence co
JOIN observation_period op ON co.person_id = op.person_id
WHERE condition_concept_id = 201826
  AND DATEDIFF(year, birth_datetime, condition_start_date) >= 18
  AND DATEDIFF(day, observation_period_start_date, condition_start_date) >= 365
```

### Why this matters

- **Reproducibility**: The SQL precisely defines who's in the cohort
- **Portability**: Run the same definition at any OHDSI site
- **Transparency**: Review and audit the exact criteria

### Exercise: Build your own

In ATLAS demo:

1. Click **New Cohort Definition**
2. Name it "Practice - [Your Name]"
3. Add an initial event (try a condition)
4. Add an inclusion criterion
5. Export and examine the SQL

---

## Clean up

```{r}
#| eval: false
disconnect(connection)
```

---

## Summary

Today we covered:

- **Git:** Create repos, stage, commit, push/pull
- **CDM Schema:** Table structure, relationships via person_id and concept_id
- **Eunomia:** Local CDM practice, SQL queries
- **SqlRender:** Dialect translation for multi-site compatibility
- **Patient Odyssey:** Following one patient through connected tables
- **ATLAS:** Visual cohort building to SQL generation

### Key takeaways

1. **Git** tracks your code history - use it for every project
2. **Remote databases** solve scale, security, and collaboration problems
3. **OHDSI's SQL tools** let you write once, run anywhere
4. The **CDM connects** through `person_id` and `concept_id`
5. **ATLAS** translates visual cohort definitions to executable SQL

### Resources

- Git tutorial: <https://git-scm.com/book/en/v2>
- OMOP CDM docs: <https://ohdsi.github.io/CommonDataModel/>
- SqlRender: <https://ohdsi.github.io/SqlRender/>
- ATLAS demo: <https://atlas-demo.ohdsi.org>
- The Book of OHDSI: <https://ohdsi.github.io/TheBookOfOhdsi/>

### Next session

We'll dive deeper into cohort definitions and start working with HADES packages for analysis.
